### 2022-03-02 paper/localpsf.tex
writing

Approximation \eqref{eq:translate1} follows from the assumption that $\Phi$ is smooth. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization of $\Phi$.
% of the form
%\begin{equation}
%	\label{eq:low_rank_interpolation}
%	\Phi(y,x) \approx \sum_{i=1}^r w_i(x) \phi_{x_i}(y)
%\end{equation}
%where $\phi_{x_i}$ are known impulse responses at a collection of points $x_i$ scattered throughout $\Omega$, and $w_i$ are smooth weighting functions that interpolate the $\phi_{x_i}$. 
%In problems of practical interest, \eqref{eq:translate1} typically holds \emph{asymptotically} in the limit as $x'$ approaches $x$. 
However, even if $\Phi$ is smooth (which is typically the case asymptotically) the relevant Lipschitz constant for continuity, and hence the required rank in the low rank approximation, can be large, which makes algorithms based on assumption \eqref{eq:translate1} computationally expensive. This is the case, for example, for Hessians in PDE constrained inverse problems with highly informative data. As the data become more informative, typically the ``mass'' of $\Phi(y,x)$ concentrates nearer and nearer to the ``diagonal'' $y=x$, the Lipschitz constant increases, and the required rank in the approximation blows up.
%Moreover, if $\Aop$ is numerically low rank, then one would typically use more accurate and robust matrix-free methods for forming low-rank approximations, such as randomized SVD or Lanczos.

@@

We start with a finite set of candidate points $P$. To build $\pointbatch_k$, first we initialize $\pointbatch_k$ as an empty set. Then we select a candidate point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in \pointbatch_1 \cup \dots \cup \pointbatch_{k-1}} \|p - x\|.
\end{equation*}
Candidate points for the first batch are chosen arbitrarily from $P$.
Once $p$ is selected, we remove $p$ from $P$. Then we perform the following two checks:
\begin{enumerate}
	\item We check whether $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in \pointbatch_k$.
	\item We make sure that $V(p)$ is not too small, in the sense $V(p) > \tau V_\text{max}$, where $V_\text{max}$ is the largest value of $V(q)$ over all points $q$ in the initial set of candidate points, and $\tau$ is a small threshold parameter (we use $\tau=10^{-5}$).
\end{enumerate}
If $p$ passes both these checks (i.e., if $p$ is sufficiently far from the other points and $V(p)$ is not too small) then we add $p$ to $\pointbatch_k$. Otherwise we discard $p$. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

We repeat the process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots$, until the number of batches exceeds a desired threshold. In our implementation, for each batch the set of candidate points $P$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem.
%, except those points in previously chosen batches.

@@

We compute impulse responses $\impulseresponse_{x_i}$ in batches by applying $\Aop$ to a Dirac combs. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the following weighted sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$:
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i} / V(x_i).
\end{equation*}
We compute the \emph{impulse response batch}, $\eta_k$, by applying $\Aop$ to the Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*.
\end{equation}
Each point source $\delta_{x_i}$ is scaled by $1/V(x_i)$ so that the resulting scaled impulse responses within a batch are all comparable in magnitude. The support of an impulse response is estimated to be contained within an ellipsoid, but this is just an approximation. A small portion of the impulse response will reside outside of the ellipsoid. Without this scaling, the neglected portion of an impulse response corresponding to a point $x_i$ could overwhelm the impulse response corresponding to a nearby point $x_j$, when $V(x_i)$ is much larger than $V(x_j)$. We are not in danger of dividing by zero here, because our ellipsoid packing procedure from Section SEC excludes $x_i$ from consideration to be a sample point if $V(x_i)$ is smaller than a predetermined threshold.

By linearity, the impulse response batch $\combresponse_k$ may be written as follows:
\begin{equation}
	\label{eq:phi_b}
	\combresponse_k = \left(\Aop \left(\sum_{x_i \in \pointbatch_k} \delta_{x_i}/V(x_i)\right)^*\right)^* = \sum_{x_i \in \pointbatch_k} \left(\Aop \delta_{x_i}^*\right)^* / V(x_i) = \sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i} / V(x_i).
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\impulseresponse_{x_i}(z) = 
	\begin{cases}
		V(x_i)\combresponse_k(z), & z \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the impulse responses $\impulseresponse_{x_i}$ associated with every point $x_i \in \pointbatch_k$.

>>>>

Approximation \eqref{eq:translate1} follows from the assumption that $\Phi$ is smooth. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization of $\Phi$.
% of the form
%\begin{equation}
%	\label{eq:low_rank_interpolation}
%	\Phi(y,x) \approx \sum_{i=1}^r w_i(x) \phi_{x_i}(y)
%\end{equation}
%where $\phi_{x_i}$ are known impulse responses at a collection of points $x_i$ scattered throughout $\Omega$, and $w_i$ are smooth weighting functions that interpolate the $\phi_{x_i}$. 
%In problems of practical interest, \eqref{eq:translate1} typically holds \emph{asymptotically} in the limit as $x'$ approaches $x$. 
However, even if $\Phi$ is smooth (which is typically the case asymptotically) the relevant Lipschitz constant for continuity, and hence the required rank in the low rank approximation, can be large, which makes algorithms based on assumption \eqref{eq:translate1} computationally expensive. This is the case, for example, for Hessians in PDE constrained inverse problems with highly informative data. As the data become more informative, typically the ``mass'' of $\Phi(y,x)$ concentrates nearer and nearer to the ``diagonal'' $y=x$, the Lipschitz constant increases, and the required rank in the approximation increases.
%Moreover, if $\Aop$ is numerically low rank, then one would typically use more accurate and robust matrix-free methods for forming low-rank approximations, such as randomized SVD or Lanczos.

@@

We start with a finite set of candidate points $P$ and build $\pointbatch_k$ incrementally with points selected from $P$. For simplicity of explanation, here $B_k$ and $P$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in \pointbatch_1 \cup \dots \cup \pointbatch_{k-1}} \|p - x\|.
\end{equation*}
Candidate points for the first batch are chosen randomly from $P$.
Once $p$ is selected, we remove $p$ from $P$. Then we perform the following two checks:
\begin{enumerate}
	\item We check whether $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in \pointbatch_k$.
	\item We make sure that $V(p)$ is not too small, by checking whether $V(p) > \tau V_\text{max}$. Here $V_\text{max}$ is the largest value of $V(q)$ over all points $q$ in the initial set of candidate points, and $\tau$ is a small threshold parameter (we use $\tau=10^{-5}$).
\end{enumerate}
If $p$ passes both these checks (i.e., if $p$ is sufficiently far from other points in the batch, and $V(p)$ is not too small) then we add $p$ to $\pointbatch_k$. Otherwise we discard $p$. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We check whether $E_p \cap E_q = \{\}$ using the ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

We repeat the process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots$, until the number of batches exceeds a desired threshold. In our implementation, for each batch the set of candidate points $P$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previously chosen batches.

@@

We compute impulse responses $\impulseresponse_{x_i}$ in batches by applying $\Aop$ to a Dirac combs. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the following weighted sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$:
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i} / V(x_i).
\end{equation*}
We compute the \emph{impulse response batch}, $\eta_k$, by applying $\Aop$ to the Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*.
\end{equation}
By linearity, the impulse response batch $\combresponse_k$ may be written as follows:
\begin{equation}
	\label{eq:phi_b}
	\combresponse_k = \left(\Aop \left(\sum_{x_i \in \pointbatch_k} \delta_{x_i}/V(x_i)\right)^*\right)^* = \sum_{x_i \in \pointbatch_k} \left(\Aop \delta_{x_i}^*\right)^* / V(x_i) = \sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i} / V(x_i).
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\impulseresponse_{x_i}(z) = 
	\begin{cases}
		V(x_i)\combresponse_k(z), & z \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the impulse responses $\impulseresponse_{x_i}$ associated with every point $x_i \in \pointbatch_k$.

Each point source $\delta_{x_i}$ in the Dirac comb is scaled by $1/V(x_i)$ so that the resulting scaled impulse responses within $\eta_k$ are comparable in magnitude. Note that we are not in danger of dividing by zero here, because our ellipsoid packing procedure from Section SEC excludes $x_i$ from consideration as a sample point if $V(x_i)$ is smaller than a predetermined threshold. Without this scaling, the portion of $\phi_{x_i}$ outside of $E_{x_i}$, which we neglect, could overwhelm $\phi_{x_j}$ for a nearby point $x_j$ simply because $V(x_i)$ is much larger than $V(x_j)$.

%Note that the support of $\impulseresponse_{x_i}$ is estimated to be contained within the ellipsoid $E_{x_i}$, but this is just an approximation. 
%%The impulse response $\impulseresponse_{x_i}$ may be nonzero inside the ellipsoid $E_{x_j}$ associated with another point $x_j$ in the batch, and this will cause error in our recovery of $\phi_{x_j}$. 
%Without this scaling, the neglected portion of an impulse response corresponding to a point $x_i$ could overwhelm the impulse response corresponding to a nearby point $x_j$ if $V(x_i)$ is much larger than $V(x_j)$.
